\include{header}

%opening
\title{\huge{Comparison of different semiparametric model-based confidence bands for survival functions}\\
	\textcolor{black}{
       	  \vspace*{5mm} \\
	  \Huge{Master's Thesis}\\
	  \vspace*{5mm}
	  \normalsize{submitted in fulfillment of\\the requirements for the degree of\\Master of Science}\\
	  \small{written at Fachhochschule Aachen\\Fachbereich Medizintechnik und Technomathematik\\
                 under supervision of\\
                 \textbf{Prof. Gerhard Dikta}}
        }
} %% title

\author{Jennifer Groß}

\date{\normalsize{Aachen, \monthyeardate\today}}

\begin{document}

\maketitle

% START: Declaration of Independence
\cleardoublepage
\thispagestyle{empty}

\newlength{\erklStart}
\addtolength{\erklStart}{\textheight}
\addtolength{\erklStart}{-40ex}
\vspace*{50ex}

\noindent I hereby declare in lieu of an oath that I have completed the present Master’s thesis independently and without illegitimate assistance from third parties. 
\\I have use no other than the specified sources and aids. In case that the thesis is additionally submitted in an electronic format, I declare that the written and electronic versions are fully identical. 
\\The thesis has not been submitted to any examination body in this, or similar, form. \\

\noindent \hspace{5em}
\begin{minipage}{20em}
	\vspace*{20ex}
	Jennifer Groß\\
	Aachen, \today
\end{minipage}
% END: Declaration of Independence

\input{chapters/abstract}

\tableofcontents

\input{chapters/introduction}

\input{chapters/theoretical_background}

\input{chapters/simulation_studies}

\input{chapters/results}

\input{chapters/summary}

\begin{appendix}

%% Bildverzeichnis
\listoffigures

%% Tabellenverzeichnsi
\listoftables

%% automatisch erzeugtes Literaturverzeichnis
\bibliographystyle{geralpha}  % mit Kürzel
%\bibliographystyle{gerunsrt}  % nummeriert
\bibliography{cites}


\end{appendix}

\clearpage
\thispagestyle{empty}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\noindent\textsf{\Huge{\textbf{\textcolor{chapter}{Acknowledgements}}}}
\vspace*{2ex}\\

\noindent I want to thank \emph{Prof. Gerhard Dikta} for supporting me on my way to this master's thesis.

Also I want to thank my friends \emph{Olaf Leidinger} and \emph{Diana Kreutz-Kielwein} for their support and motivation effort.

My boyfriend \emph{Justus Schwan} always had my back and stood beside me through this long journey. Thank you for that.


\end{document}
